---
title: "Including health impacts in transport modelling and economic appraisal-ITHIM Melbourne"
author: "Belen Zapata-Diomedi, belen.zapata-diomedi@rmit.edu.au, Alan Both, alan.both@rmit.edu.au"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    df_print: paged
    toc : true
    number_sections: true
  pdf_document: 
    df_print: paged
    keep_tex: yes
    number_sections: yes
    toc: yes
biblio-style: apsr
header-includes:
- \usepackage{graphicx}
- \usepackage{subfig}
- \usepackage{floatrow}
- \floatsetup{capposition=top}
- \floatplacement{figure}{H}
- \captionsetup{position=above, aboveskip=0pt, belowskip=0pt}
bibliography: DoT.bib
abstract: Short tehcnical abstract
---
```{r setup, include=FALSE}
## Global options
knitr::opts_chunk$set(echo = TRUE, width = 75)
options(knitr.table.format = 'latex')
options(tinytex.verbose = TRUE)

```

```{r Libraries, echo = FALSE, message = FALSE, warning = FALSE}
library(knitr)
library(kableExtra)
library(plyr)
library(lubridate)
library(ggplot2)
library(ggjoy)
library(ggpubr)
library(plm)
library(PerformanceAnalytics)
library(pander)
library(dplyr)
library(tidyr)
library(tibble) 
library(knitr)
library(readxl)
library(stringr)
library(Hmisc)
library(DiagrammeR)
library(tinytex)
options(scipen = 999)
```


\newpage

# Introduction

In this document we present the methods and data sources for the development of the quantitative health impact assessment model for the impacts of physical activity scenarios on population health. We developed two models: 1) Melbourne model and 2) Australia model. The Melbourne model quantifies physial activity related health impacts of changes in transport, and the Australia model quantifies health impacts of changes in physical activity directly, not informed by changes in transport. 
In what follows, Section 2 explains the quantitative health impact model, including methods, data inputs and modelling assumptions. Section 3 explains an application of the Melbourne model to an active transport scenario developed in collaboration with the Victorian Department of Transport and its results. Section 4 demonstrates the use of the web interface throught the Australian Urban Observatory. 

# Quantitative health impact assessment model overview

The model analytical framework is depicted in Figure 1. The model consists of three main sections: 1) Scenarios; 2) Potential Impact Fractions and Risk calculations; and 3) Proportional multi-state life table (PMSLT). The simulation starts with a change in physical activity in the population of interest. For the Melbourne model changes in physical activity are informed by the transport scenarios and in the Australian model such changes are made direclty to physical activity. A change in physical activity, combined with the risk of diseases associated to it, is combined in the potential impact fraction (PIF) calculations to then modify incidence in the disease process (Proportional Multi-State Life Table). From each of the diseases process, data on changes in mortality and prevalence of diseases is collected to then modify overall mortality and prevalent years years lived with disability in the general life table. Outcomes by age (5-years) and sex cohort include: change in the number of incidence cases, prevalent years years lived with disability, life years, health-adjusted life years, life expectancy and health adjusted-life expectancy.

\blandscape
```{r, echo = FALSE, fig.cap = 'Figure 1: Analytical Framework', out.height= "200%", out.width="100%"} 
### This graph needs some work! basics are here

analytical_framework <- grViz("digraph{
         
                   graph[rankdir = LR]
                     
                      # graph, node, and edge definitions
                      graph [compound = true, nodesep = .25, ranksep = 2.25,
                      color = black, overlap = true, height = 200]
                      
                      node [fontname = Calibri, fontcolor = black,
                      shape = rectangle, fixedsize = false, width = 3.5,
                      color = DarkGray]

                      edge [color = DarkGray, arrowhead = none, arrowtail = none, tailport= E]

                         subgraph cluster_0 {
                          graph[shape = rectangle]
                          style = rounded
                          bgcolor = WhiteSmoke
                          fontsize = 30
                          face = bold
                          fontname = Calibri
                          color = Gray
                         
                          label = 'Scenarios'
  
                            node[shape = rectangle, color = Black, margin = 0.25, fontsize = 32, style = dotted]
                            0[label = 'Physical activity']
                            1[label = 'Private car use']
                            2[label = 'Cycling']
                            3[label = 'Walking']
                            4[label = 'Public Transport'] 
                            
                            {rank = same; 0;2;3;4}
                          
       }  
                          
                            subgraph cluster_1 {
                          graph[shape = rectangle]
                          style = rounded
                          bgcolor = WhiteSmoke
                          fontsize = 30
                          fontname = Calibri
                          color = Gray
                          label = 'Potential impact fraction and risk'
                          
                                              subgraph cluster_10 {
                          graph[shape = rectangle]
                          style = rounded
                          bgcolor = WhiteSmoke
                          fontsize = 30
                          fontname = Calibri
                          color = Gray
                          label = ''
                            
                             node[shape = rectangle, color = Black, margin = 0.25, fontsize = 32, style = dotted]
                            # 5[label = 'Ambient PM2.5']
                            # 6[label = 'On road PM2.5']
                             7[label = 'Physical activity']
                            # 8[label = 'Road trauma']
                          }   
                          
                          
                          subgraph cluster_11 {
                          graph[shape = rectangle]
                          style = rounded
                          bgcolor = WhiteSmoke
                          fontsize = 30
                          fontname = Calibri
                          color = Gray
                          label = 'Relative risks'
                            
                            node[shape = rectangle, color = Black, margin = 0.25, fontsize = 32, style = dotted]
                            9[label = 'Ischemic heart disease']
                            10[label = 'Stroke']
                            # 11[label = 'COPD']
                            # 12[label = 'Lower respiratory infection']
                            13[label = 'Diabetes mellitus type 2']
                            14[label = 'Breast cancer']
                            15[label = 'Colon cancer']
                            16[label = 'Uterine cancer']
                          }
                            }
                            
                          subgraph cluster_2 {
                          graph[shape = rectangle]
                          style = rounded
                          bgcolor = WhiteSmoke
                          fontsize = 30
                          fontname = Calibri
                          color = Gray
                          label = 'Proportional multi-state life table'
                             
                            node[shape = rectangle, color = Black, margin = 0.25, fontsize = 32, style = dotted]
                            17[label = 'Disease processes']
                            # 18[label = 'Injuries process']
                            19[label = 'Life table']
                            
                          }  
                                subgraph cluster_3 {
                           graph[shape = rectangle]
                          style = rounded
                          bgcolor = WhiteSmoke
                          fontsize = 30
                          fontname = Calibri
                          color = Gray
                          label = 'Outcomes'
                            
                            node[shape = rectangle, color = Black, margin = 0.25, fontsize = 32, style = dotted]
                            20[label = 'Incidence cases']
                            21[label = 'Disease deaths']
                            # 22[label = 'Road trauma pYLDs']
                            # 23[label = 'Road trauma deaths']
                            
                            [label = 'Life years']
                            25[label = 'Health-adjusted life years']
                            26[label = 'Life expectancy']
                            27[label = 'Health-adjusted life expectancy']
                           } 
                            
                            edge[color = Black, arrowhead = normal, arrowsize = 1.25]
                          0 -> 7
                          1 -> {2 3 4} # 5 6 8
                          2 -> 7 # 5 6 8
                          3 -> 7 # 5 6 8
                          4 -> 3 # 5 6 8
                          
                          
                          edge[color = DarkGray, arrowhead = normal, arrowsize = 1.25]
                          # {5 6} -> {9 10 11 12}
                          7 -> {9 10 13 14 15 16}
                          {9 10 13 14 15 16} -> 17  # 11 12
                          # 8 -> 18
                          # {17 18} -> 19
                          17 -> {20 21}
                          # 18 -> {22 23}
                          19 -> {
                          25 26 27}
                          17 -> 19
                     }")

analytical_framework

```
\elandscape

## Scenarios

Table 1 outlines user inputs for each of the models: Melbourne model and Australian model. 

```{r, echo=FALSE, include=TRUE}

scenarios <- tribble(
~"User Input", ~"Melbourne Model", ~"Australia model",
"Age", "All ages, 18-19, 20-
, 25-29, 30-34, 35-39, 40-44, 45-49, 50-54, 55-59, 60-64, 65-69, 70-74, 75-79, 80-84, 85-89, 90-94, 95-98", "All ages, 15-19, 20-
, 25-29, 30-34, 35-39, 40-44, 45-49, 50-54, 55-59, 60-64, 65-69, 70-74, 75-79, 80-84, 85-89, 90-94, 95-99",
"Sex", "Female, Male, Both",  "Female, Male, Both",
"Physical activity minutes", "NA", "Number: 0 to xx",
"Physical activity distance", "NA", "Number: 0 to xx", 
"Replace mode", "Walking or cycling", "N/A", 
"Replace distance", "Number: 0 to xx", "N/A", 
"Trip purpose", "social, buy something, work related, pick-up or drop-off someone, personal business, unknown purpose (at start of day),recreational, pick-up or deliver something, accompany someone,education, other purpose, at or go home, change mode, not stated", "N/A", 
"Day of the week", "Weekday, Weekend day, both", "N/A"
)
kable(scenarios, booktabs = T, caption = "Table 1: Inputs PA-health impacts model online interface", longtable = TRUE) %>%
column_spec(1:3, width = "15em") %>%
  row_spec(0, bold = T) %>%
  kable_styling()
```

## Potential Impact Fraction

The potential impact fraction (PIF) measures the proportional change in the burden of disease attributable to a change in a related risk factor [@RN19]. We used PIFs to calculate the average change in incidence of diseases (Figure 2) in the scenarios' diseases' processes. PIFs were derived for: ishcemic heart disease, stroke, type 2 diabetes, lung cancer, colon cancer, breast cancer and uterine cancer. The methods developed in the ITHIM-R (https://github.com/ITHIM/ITHIM-R) project were used here to calculate individual level RRs for BAU and scenario which where then used to calculate individual level PIFs (\ref{eq1}).From individual level PIFs, age group and sex average PIFs were calculated to modify incidence in the age and sex cohort disease processes' (Figure 3). Specifically the function to generate relative risks at the individual level for BAU and sceanarios (gen_pa_rr). Inputs of gen_pa_rr are individual marginal metabolics of taks per hour (mMET-hours) per week. METs are metabolic equivalent of tasks and measure the energy expenditure required to carry out the physical activity, expressed as a multiple of the resting metabolic rate. Marginal METs only consider energy expended above resting metabolic rate.[@RN21]  All diseases have a threshold beyond which there are no benefits from increases in mMETs-hours, except for diabetes (35 mMET-hours for: colon, breatst and uterine cancers, ishcemic heart disease and stroke; 10 for lung cancer and 32 for stroke) (https://github.com/ITHIM/ITHIM-R). Table 2 oulines RRs mean, lower and upper bounds for each of the modelled diseases from which the samples were taken following normal distributions truncated to zero (https://www.overleaf.com/project/5bc08a8caba5c52e6060b614). Individual level RRs for BAU and sceanrio and then modelled from the quantile (CHECKING with JAMES). mMETs-hours were derived by summing up the product of physical activity duration and intensity [@RN23].
Below, calculations for mMET-hours for the Melbourne and Australian model are explained in turn and MET values with associated uncertainty (Table 4).

Relative risks sources: https://shiny.mrc-epid.cam.ac.uk/meta-analyses-physical-activity/
$$
\begin{align}
 PIF_d= \frac {RR_{dBAU} - RR_{dSCENRIO}} {RR_{dBAU}}
\end{align}
$$

### Melbourne model

We created a matched population to derive individual mMET-hours per week based on the **Victorian Integrated Survey of Travel and Activity 2017-2018** (VISTA 2017-18 @RN20) and the **National Health Survey, 2017-18, Basic CURF** (NHS, @RN22) physical activity data collected for surveillance. PA data from the NHS were allocated to each VISTA individual. MET-hours per individual were derived from the sum of the product of hours spent weekly doing transport walking, transport cycling, vigorous PA for leisure and work, moderate PA for leisure and work and walking for leisure by their corresponding mMET-hours scores. Tranport walking and cycling were sourced from the VISTA survey and the rest from the NHS (Table 2).
We used the latest VISTA survey conducted across the period 2017-18 for Greater Melbourne and Geelong. Randomly selected households were asked to complete the household form and each individual to complete a travel diary for a single specified day (or adult for children). The survey aims to capture average travel behaviour. The Greater Melbourne data includes 3,543 households, 7,117 individuals and 23,166 trips. Travel data was collected for all households' persons above the age of 5. We limited the analysis to individuals over the age of 16. The travel diary asked respondents about travel and activities on a particular travel day, specifically: 
1) all travel over the whole day, from 4:00 am on the travel day until 4:00 am the next day,
2) all trips to be included, even short trips (like walking to lunch or going for a jog), and
3) includes people who do not travel on the travel survey day. 

From the **VISTA** survey we obtained time and distance spent by mode of travel per week. We created time and distance variables per person by adding up all stages time and distance by mode in a trip. A trip may have up to 9 stages. The data is for one day, and the survey indicates whether travel occured on a weekday or weekend day. We derived a week total time and distance by mode by multiplying by 5 for weekdays and 2 for weekend days.


##### ADD VISTA STATS HERE

The **NHS** is a representative sample of Australian adults (>=15 years of age). The data were collected by the Australian Bureau of Statistics and published as the National Health Survey, 2017-18, Basic-CURF [@RN22]. A sample of 21,315 adults was available. Survey respondents were asked about the time they spent doing exercise (for leisure or walking for transport) and workplace physical activity in the last week. Table 3 outlines types of physical activity data collected in the NHS and table 4 all physical activity variables measured in a time unit (minutes). Table 5 depics descriptive statistics (proportion by age and sex, population, sample, means, sd and quantiles for minutes spent in leisure and transport related PA)
```{r, echo=FALSE, include=TRUE}

pa_NHS_collected_data <- tribble(
~"Type of activity", ~"Definition",
"Walking for fitness, recreation and sport", "Must have carried out for at least 10 minutes continuously", 
"Walking to get from place to place", "1) Excludes activities such as walking around a shopping centre as people tend to frequently pause while shopping. 2) Must have carried out for at least 10 minutes continuously.", 
"Moderate intensity exercise", "1) Comprises activities that caused a moderate increase in the heart rate or
breathing of the respondent (e.g. gentle swimming, social tennis, golf). 2) Excludes any previously identified walking as well as household chores, gardening or yard work, and any activity carried out as part of a job", 
"Vigorous intensity exercise", "1) Comprises activities that caused a large increase in the respondent's
heart rate or breathing (e.g. jogging, cycling, aerobics, competitive tennis). 2) Excludes any previously identified walking as well as household chores, gardening or yard work, and any activity carried out as part of a job",
"Strength or toning exercises", "1) Activities designed to increase muscle strength or tone, such as lifting
weights, resistance training, pull-ups, push-ups, or sit ups. 2) Includes any Strengthening and toning exercises activities already mentioned",
"Workplace physical activity", "1) Usual physical activity while at work on a typical work day. 2) Must have carried out for at least 10 minutes continuously. 3) Moderate activity in the workplace was defined as activity that caused a moderate increase in the heart rate or breathing of the respondent (e.g. brisk walking or carrying light lights). 4) Vigorous activity in the workplace was defined as activity that caused a large increase in heart rate or breathing (e.g. carrying or lifting heavy loads, digging or construction work)")

kable(pa_NHS_collected_data, booktabs = T, caption = "Table 3: NHS PA variables", longtable = TRUE) %>%
column_spec(1:1, width = "15em") %>%
  row_spec(0, bold = T) %>%
  kable_styling()
```


```{r, echo=FALSE, include=TRUE}

pa_NHS_variables <- tribble(
~"Variable name", ~"Descriptive table name",
"EXLWTIME: Total minutes undertaken exercise for fitness, recreation, sport or transport in last week
", "MeanTotal, QuantileTotal", 
"EXLWTIM2	Total minutes did exercise or walked for transport in last week (vig x 2)", " MeanTotal2, QuantileTotal2", 
"EXFSRMIN	Total minutes walked for fitness, recreation or sport in last week", "MeanWalkRec, QuantileWalkRec",
"EXTRAMIN	Total minutes spent walking for transport in last week", "MeanWalkTrans, QuantileWalkTrans",
"EXWLKTME	Total minutes spent walking for exercise and transport last week", "MeanWalkTotal, QuantileWalkTotal",
"EXLWMMIN	Total minutes undertaken moderate exercise last week","MeanModExe, QuantileModExe",
"EXLWVMIN	Total minutes undertaken vigorous exercise last week", "MeanVigExe, QuantileVigExe",
"MODMINS	Total minutes undertaken all moderate physical activity last week", "MeanTotalMod, QuantileTotalMod", 
"VIGMINS	Total minutes undertaken all vigorous physical activity last week", "MeanTotalVig, QuantileTotalVig", 
"PAMINS	Total minutes undertaken all physical activity last week", "MeanTotalPA, QuantileTotalPA", 
"PAMINS2	Total minutes undertaken all physical activity last week (vig x 2)", " MeanTotalPA2, QuantileTotalPA2")

kable(pa_NHS_variables, booktabs = T, caption = "Table 4: NHS PA variables", longtable = TRUE) %>%
  kable_styling()
```


```{r, echo=FALSE, include=TRUE}

minutes_NHS_stats_non_work <- readRDS("SuppDocs/Tables/Minutes.Stats.rds") %>%
  dplyr::rename(Sex = SEX, `Age group` = age_group) %>%
  dplyr::select(-c("Proportion_se", "Population_se"))
kable(minutes_NHS_stats_non_work  , booktabs = T, caption = "Table 5: Summary statistics NHS PA leisure and transport variables (PA variables are in hours per week", longtable = TRUE) %>%
  kable_styling()
```

```{r, echo=FALSE, include=TRUE}

minutes_NHS_stats_work <- readRDS("SuppDocs/Tables/MinutesWork.Stats.rds") %>%
  dplyr::rename(Sex = SEX, `Age group` = age_group) %>%
  dplyr::select(-c("Proportion_se", "Population_se"))
kable(minutes_NHS_stats_work , booktabs = T, caption = "Table 6: Summary statistics NHS work PA variables (PA variables are in hours per week", longtable = TRUE) %>%
  kable_styling()
```

Minutes spent doing moderate physical activity for leisure and work, vigorous physical activity for leisure and work, walking for leisure during the week were used in this analysis. The three categories are mutually exclusive [@RN24] and were translated to hours per week. 
Total mMETs per person for the BAU scenario were derived by translating minutes spent in each of the PA categories per week to hours and then multiplying by the MET scores [@RN21] minus one. Tables 3 outlines all types of PA activities included, by source, and correspoinding assigned MET scores. VISTA individuals were matched to NHS individuals based on age, sex, socio-economic status, whether they walk for transport and work status (see Table 3).  
  
```{r, echo=FALSE, include=TRUE}

matched_population <- tribble(
~"Matching variable", ~"VISTA", ~"NHS",
"Age", "AGE: single year of age recoded to groups to match NHS's groups", "AGEB: 0 - 4 years to 80 - 84 years in five year age groups and 85 years and over years",
"Sex", "SEX: Male/Female", "SEX: Male/Female",
"Socio-economic Status", "VISTA data did not provida household SES data. Australian postcode SEIFA - Index of Relative Socio-economic Disadvantage - 2016 - SA1 - Deciles Postal Area code  (POA) [@RN23] was matched to individuals' household POA to assign them an area level IRSD", "SA1SF2DN: SEIFA - Index of Relative Socio-economic Disadvantage - 2016 - SA1 - Deciles - National",
"Walk for transport", "Whether VISTA person had transport walking trips" ,"EXLWKTNO: Number of times walked for 10 minutes or more for transport in the last week",
"Work status","ANYWORK: employed, unemployed", "LFSBC (Labour force status): 0. Not applicable (recoded to unemployed), 1. Employed, 2. Unemployed, 3. Not in the labour force (recoded to unemployed)"
)
kable(matched_population, booktabs = T, caption = "Table 1: Matched poplation", longtable = TRUE) %>%
column_spec(1:3, width = "15em") %>%
  row_spec(0, bold = T) %>%
  kable_styling()
```

### Australian model

The main difference between the Melbourne and Australian model is the calculation of mMETs per person. In the Australian model, only the NHS data was used, above described. Total mMET-hours were derived from all four variables form the NHS described in Table 3 (moderate, vigourus and all walking (for transport and recreation)). 


```{r, echo=FALSE, include=TRUE}
mets <- tribble(
~"Physical activity", ~"Survey", ~"mMETs", ~"Source", ~"Model", 
"Walking for transport", "VISTA 2017-18", "2.53 (1.1)", "ITHIM-R, check with James", "Melbourne", 
"Cycling for transport", "VISTA 2017-18", "4.63 (1.2)", "ITHIM-R, check with James", "Melbourne",
"Walk for leisure", "NHS 2017-18", "2.53 (1.1)", "Compedium of PA [@RN21]", "Melbourne and Australia",
"Walk for transport", "NHS 2017-18", "2.5", "Compedium of PA [@RN21]", "Australia",
"Total moderate excersice and work", "NHS 2017-18", "3.5", "Compedium of PA [@RN21]", "Australia",
"Total vigorous excersice and work", "NHS 2017-18", "7", "Compedium of PA [@RN21]", "Australia"
 )
kable(mets, booktabs = T, caption = "Table 3: Marginal METs", longtable = TRUE) %>%
column_spec(1:4, width = "15em") %>%
  row_spec(0, bold = T) %>%
  kable_styling()
```



### Some graphs here for distribution of physical actiity (mets) comparing melbourne modle and Australia models

## Proportional multi-state life table

The PMSLT is a population level model (macro) approach to simulate health (and economic) implications of changes in exposure to health risk factors (e.g. physical inactivity, air pollution and diet). The PMSLT has been widely used to simulate outcomes for population level interventions for the reduction risks for chronic diseases. Originally developed by [@RN17]. Figure 2 is an schematic representation of the PMSLT The mechanisms of the the PMSLT and associated diseases and injuries processes are e along with input data (Table 1). 

\blandscape
```{r, echo = FALSE, fig.cap = 'Figure 1: Analytical Framework', out.height= "200%", out.width="100%"} 
### This graph needs some work! basics are here

PMSLT_graph <- grViz("digraph{
         
                   graph[rankdir = LR]
                     
                      # graph, node, and edge definitions
                      graph [compound = true, nodesep = .25, ranksep = 2.25,
                      color = black, overlap = true]
                      
                      node [fontname = Calibri, fontcolor = black,
                      shape = rectangle, fixedsize = false, width = 2,
                      color = DarkGray, fontsize = 30]
                      
                      edge [color = DarkGray, arrowhead = none, arrowtail = none, tailport= E]
                      
                      
                      ### Life tables subgroup
                      #### Graph cluster with BAU and Intervention Life table
                      
                          subgraph cluster_0 {
                           graph[shape = rectangle, rankdir=LR]
                           style = rounded
                           bgcolor = WhiteSmoke
                           fontsize = 30
                           face = bold
                           fontname = Calibri
                           color = Gray
                            label = '' 
                     
                          
                      ##### Life table graph BAU    
                          subgraph cluster_1 {
                          graph[rankdir = LR]
                          graph[shape = rectangle]
                          style = rounded
                          bgcolor = WhiteSmoke
                          fontsize = 30
                          face = bold
                          fontname = Calibri
                          color = Gray
                         
                          label = 'Life Table BAU'
  
                            node[shape = rectangle, color = Black, margin = 0.25, fontsize = 32, style = dotted, group = right]
                            edge[color = DarkGray, arrowhead = normal, arrowsize = 1.25]
                            1[label = 'mx']
                            2[label = 'qx']
                            3[label = 'lx']
                            4[label = 'Lx']
                            5[label = 'pyldx']
                            6[label = 'Lwx']
                            1 -> 2 ->3 ->4;
                            4 -> 6;
                            5 -> 6
                         }
                            

                          ##### Life table graph Intervention 
                          subgraph cluster_2 {
                          graph[shape = rectangle]
                          style = rounded
                          bgcolor = WhiteSmoke
                          fontsize = 30
                          face = bold
                          fontname = Calibri
                          color = Gray
                         
                          label = 'Life Table intervention'
  
                            node[shape = rectangle, color = Black, margin = 0.25, fontsize = 32, style = dotted, group = right]
                            edge[color = DarkGray, arrowhead = normal, arrowsize = 1.25]
                            7[label = 'mx']
                            8[label = 'qx']
                            9[label = 'lx']
                            10[label = 'Lx']
                            11[label = 'pyldx']
                            12[label = 'Lwx']
                            7 -> 8 ->9 ->10;
                            11 -> 12[ dir = back ];
                            10 -> 12
                     
                              }}
            
                        
                        ### Disease process subgroup    
                          subgraph cluster_3 {
                           graph[shape = rectangle]
                           style = rounded
                           bgcolor = WhiteSmoke
                           fontsize = 30
                           face = bold
                           fontname = Calibri
                           color = Gray
                            label = '' 
                            
                            
                            
                          subgraph cluster_4 {
                          graph[shape = rectangle]
                          style = rounded
                          bgcolor = WhiteSmoke
                          fontsize = 30
                          fontname = Calibri
                          color = Gray
                          label = 'Disease process BAU'
                            
                            node[shape = rectangle, color = Black, margin = 0.25, fontsize = 32, style = dotted, group = left]
                            edge[color = DarkGray, arrowhead = normal, arrowsize = 1.25]
                            13[label = 'ix']
                            14[label = 'px']
                            15[label = 'mx']
                            16[label = 'w']
                            17[label = 'pylds']
                            13 -> 14 -> 15;
                            14 -> 17; 
                            16 -> 17
                            
                          }

                          subgraph cluster_5 {
                          graph[shape = rectangle]
                          style = rounded
                          bgcolor = WhiteSmoke
                          fontsize = 30
                          fontname = Calibri
                          color = Gray
                          label = 'Disease process intervention'
                          
                            node[shape = rectangle, color = Black, margin = 0.25, fontsize = 32, style = dotted, group = left]
                            edge[color = DarkGray, arrowhead = normal, arrowsize = 1.25]
                            18[label = 'ix']
                            19[label = 'px']
                            20[label = 'mx']
                            21[label = 'w']
                            22[label = 'pylds']
                            18 -> 19 -> 20;
                            19 -> 22;
                            21 -> 22; 
                            
                          }
                          subgraph cluster_6 {
                          graph[shape = rectangle]
                          style = rounded
                          bgcolor = WhiteSmoke
                          fontsize = 30
                          fontname = Calibri
                          color = Gray
                          label = 'Injury process baseline'
                          
                            node[shape = rectangle, color = Black, margin = 0.25, fontsize = 32, style = dotted, group = left]
                            edge[color = DarkGray, arrowhead = normal, arrowsize = 1.25]
                            23[label = 'mx']
                            
                            [label = 'pyldx']
                            
                          }
                          
                          subgraph cluster_8 {
                          graph[shape = rectangle]
                          style = rounded
                          bgcolor = WhiteSmoke
                          fontsize = 30
                          fontname = Calibri
                          color = Gray
                          label = 'Injury process intervention'
                          
                            node[shape = rectangle, color = Black, margin = 0.25, fontsize = 32, style = dotted, group = left]
                            edge[color = DarkGray, arrowhead = normal, arrowsize = 1.25]
                            25[label = 'mx']
                            26[label = 'pyldx']
                            
                          }}
                          
                          subgraph cluster_7 {
                           graph[shape = rectangle, rankdir=LR]
                           style = rounded
                           bgcolor = WhiteSmoke
                           fontsize = 30
                           face = bold
                           fontname = Calibri
                           color = Gray
                     
                           label = 'Difference intervention and BAU'
                     
                             node[shape = rectangle, color = Black, margin = 0.25, fontsize = 32, style = dotted, group = right]
                             edge[color = DarkGray, arrowhead = normal, arrowsize = 1.25]
                             27[label = 'mx']
                             28[label = 'pyldx']
                
                          }

  
                   
                    edge[color = DarkGray, arrowhead = normal, arrowsize = 1.25, style= dotted]
                   
                   
                    {15 20 23 25} -> 27 
                    {17 22 
                    26} -> 28 
                    27 ->7
                    28 -> 11
                     
                     }")

PMSLT_graph
```
\elandscape


## Data inputs


Data inputs for the proportional multi-state life table model (PMSLT) include inputs for: general life table (population, mortality rates and prevalent years lived with disability rates for all causes (YLD)), diseases life tables (incidence and case fatality) and injuries process (mortality rates and years lived with disability rates). Here we prepared inputs for the Melbourne population, when local inputs were available, these were used, otherwise, we used state or nationally representative inputs. Table 1 describes all data inputs for the MSLT. Each of the MSLT components and related code are explained below Table 1.

```{r, echo=FALSE, include=TRUE}

pmslt_inputs <- tribble(
~Component, ~"Data needs", ~"Input data", ~"Data processing",
"General life table", "Population numbers and mortality rates", "Population Greater Melbourne [@RN2] by 5-year age groups and sex and death rates by single year of age and sex [@RN11]", "Population data is from the ABS 2016 Census. Mortality rates are for Victoria given that future projections for mortality were available for Vicotoria and not for Melbourne. Deaths rates projections were available by 1-year age intervals and sex from 2017 (baseline to 2066)", 
"General life table", "All-cause YLDs", "Global Burden of Disease (GBD) study by 5-year age groups and sex [@RN8]", "GBD data is in five-year age groups, interpolation to derive one-year age groups", 
"Disease life table", "Disability weights, incidence and case fatality", "Disability weights were derived from disease specific YLDs [@RN8] and prevalence [@RN8]. Incidence and case fatality were derived using DISMOD II  and ", "Table 2 outlines DISMOD II data processing and data inputs. Disability weights were calculated by dividing disease specific YLDs by prevalence and adjusted by all cause YLDs",
"Diseases trends", "Future trends to apply to incidence and case fatality", "Trends cancers: incidence [@RN3], mortality [@RN4]. Trends for cardiovascular diseases, diabetes and chronic obstructive pulmonary disease were not available. These were derived from past data [@RN5; @RN6; @RN7]", "Explained in sections Trends"
)
kable(pmslt_inputs, booktabs = T, caption = " Table 2: PMSLT inputs", longtable = TRUE) %>%

  column_spec(1:2, width = "7em") %>%
  column_spec(3:4, width = "15em") %>%
  row_spec(0, bold = T)  %>%
  kable_styling()
```

### General life table: population numbers and mortality rates

As outlined in Table 1, the PMSLT was parametrised with population data for Melbourne (**will change to the population impacted by the intervention**) and mortality rates for Victoria. Below, Table 2 summarizes population data by age and sex and Graphs 1 and 2 are deaths rates for each age and sex cohort. The function (*GetDeathsRates*) is used to sort out original ABS data to a suitable input format for this model.

```{r,echo=FALSE, message=FALSE}

#  population <- read.csv("Data/Processed/population_melbourne_abs.csv", as.is=T,
#                          fileEncoding="UTF-8-BOM") %>%
#     # population should be numeric
#     mutate(population = as.numeric(gsub(",", "", population))) %>%
#     # some ages have space at the end
#     mutate(age = gsub(" ", "", age)) %>%
#     mutate(age = gsub("-", " to ", age)) %>%
#     mutate(sex = tolower(sex)) %>% 
#     rename(`Age cohort` = age, Sex = sex, Population = population) %>%
#     arrange (Sex) %>%
#     pivot_wider(names_from=c("Sex"),
#                 values_from=Population)
# 
# kable(population, booktabs = T, caption = "Melbourne population in 2017") %>%
#  kable_styling(latex_options = c("hold position", "repeat_header"),
#                full_width = F)%>%
#   column_spec(2, width = "15em") %>%
#   column_spec(3, width = "15em") %>%
#   row_spec(0, bold = T) %>%
#   kable_styling()
# 
# 
# # Deaths
# source("Scripts/data_prep/death_rates_prep.R")
# death_rates <- GetDeathsRates(
#   deaths="Data/Population and deaths/projections.xls"
# )
# 
# write.csv(death_rates[[1]], "Data/Processed/deaths_rates_males.csv", row.names=F, quote=T)
# write.csv(death_rates[[2]], "Data/Processed/deaths_rates_females.csv", row.names=F, quote=T)

```

```{r,echo=FALSE, message=FALSE}
# 
# #### NEED TO FIX TITLES
# ## Plot to check data
# 
# age_cohort =seq(from = 17, to = 100, by = 5)
# age_cohort_names <- c("15-19", "20-
#                       ", "25-29", "30-34", "35-39", "40-44", "45-49", "50-54", "55-59", "60-64", "65-69", "70-74", "75-79", "80-84", "85-89", "90-94", "95-100")
# 
# data_males <- list()
# index <- 1
#   for (a in age_cohort) {
# 
#     data_males[[index]] <-  as.data.frame(death_rates[[1]]) %>% filter(age_cohort == a)
# 
#     # print(plotdf1[[index]])
#   index <- index +1
#   
#   }
# 
# p_males <- ggplot(bind_rows(data_males, .id="Cohort"), aes(age, rate)) +
#   geom_line()
# p_males + facet_wrap(~age_cohort, scales = "free") +
#    facet_wrap(~ age_cohort, scales = "free",ncol=3, 
#       # add a labeller here:
#       labeller = as_labeller(setNames(age_cohort_names, sort(unique(age_cohort))))) +
#   ggtitle("Deaths rates for male cohort from baseline year (2017) up to age 100") +
#   xlab("Age cohort") +
#   ylab("Mortality rate") +
#   theme(legend.title=element_blank()) +
#   theme_classic()
# 
# 
# 
# data_females <- list()
# index <- 1
#   for (a in age_cohort) {
# 
#     data_females[[index]] <-  as.data.frame(death_rates[[2]]) %>% filter(age_cohort == a)
# 
#     # print(plotdf1[[index]])
#   index <- index +1
#   
#   }
# 
# p_females <- ggplot(bind_rows(data_females, .id="Cohort"), aes(age, rate)) +
#   geom_line()
# p_females + facet_wrap(~age_cohort, scales = "free")  +
#    facet_wrap(~ age_cohort, scales = "free",ncol=3, 
#       # add a labeller here:
#       labeller = as_labeller(setNames(age_cohort_names, sort(unique(age_cohort))))) +
#   ggtitle("Deaths rates for female cohort from baseline year (2017) up to age 100") +
#   xlab("Age cohort") +
#   ylab("Mortality rate") +
#   theme(legend.title=element_blank()) +
#   theme_classic()
#   
```

### General life table: All cause prevalent Years with Disability (pYLDs)

 PYLDs for all causes were obtained in 5 year age groups adn sex from the GBD study for 2017. Interpolation was used to derive rates per single year of age. The function *CalculateMSLT* was used for obtaining pYLDs for all causes. 


### Disease life table: disability weights

Disability weights (DW) are derived from disease specific prevalent years lived with disability (YLD) and disease specific prevalence by age group (5 years) and sex. Data for YLDs prevalence from GBD data for 2017. An age and sex specific-correction was introduced to counteract the effects of accumulating comorbid illnesses in the older age groups (Equation 1). Interpolation was used to derive rates per single year of age from 5-year age intervals rates with the function *CalculateMSLT*. 

\begin{equation}
\label{DW adjusted for total YLDs-YLDdPd1YLDt}
(YLDd/Pd)/(1-YLDt) = DW adjusted for total YLDs
\end{equation}


Where YLDd is the YLD mean number per age and sex for a given disease, Pd is the prevalence (as reported in GBD )  for a given disease by age and sex and YLDt is total YLD rate per age and sex. 


### Disease life table: incidence and case fatality

We used AIHW incidence and mortality cancer data for 2017 for Australia and GBD data for Australia for other diseases for the year 2017 to derive incidence and case fatality inputs for breast cancer (females), colon cancer, lung cancer, uterine cancer (females), ischemic heart disease, stroke, type 2 diabetes and chronic pulmonary disease. These diseases were selected as the modeled risk factors for these diseases are low levels of physical activity and exposure to Particulate matter of 2.5 micrometers and smaller. TO DO: injuries, depending on weather we include or not. 
We used Dismod II (free at https://www.epigear.com/index_files/dismod_ii.html) to derive internally consistent inputs for incidence and case fatality and generate missing data. For example, the GBD studies provide data for incidence, prevalence and disease mortality, however, not case fatality. For cancers we used AIHW given that the GBD data was derived assuming remission for cancers (except for long life sequelaes) (@RN10), but in our model we assumed no remission for diseases. In addition, AIHW generates data for colon cancer and lung cancer, while the GBD only provides estimates for larger disease groups (colon and rectum cancer and tracheal, bronchus, and lung cancer). Dismod II is a separate tool, therefore, here, we provide data information to obtain data inputs for Dismod II (*gbd_wider* and *calculateInputsDismodAIHW *) and the procedures used to generate the data (Table 2). 


```{r,echo=FALSE, message=FALSE, warning=FALSE}
source("Scripts/data_prep/mslt_gbd_prep.R")
options(scipen=999)
gbd_wider <- calculateGBDwider(
  gbd_location="Data/original/gbd/gbd_mslt.csv"
)
write.csv(gbd_wider, "Data/processed/mslt/gbd_wider.csv", row.names=F, quote=T)

source("Scripts/data_prep/aihw_prep.R")
AIHW <- calculateInputsDismodAIHW (
incidence_AIHW="Data/original/aihw/cancer_incidence_AIHW_with_projections.xlsx",
mortality_AIHW="Data/original/aihw/cancer_mortality_AIHW_with_projections.xlsx",
population_deaths_location="Data/original/abs/population_deaths.csv"
)


write.csv(AIHW[[1]], "Data/dismod/incidence_AIHW_dismod.csv", row.names=F, quote=T)
write.csv(AIHW[[2]], "Data/dismod/mortality_AIHW_dismod.csv", row.names=F, quote=T)
write.csv(AIHW[[3]], "Data/dismod/mortality_pop_AIHW.csv", row.names=F, quote=T)
```


```{r,echo=FALSE, message=FALSE, warning=FALSE}

dismod_process <- tribble(
~"Disease", ~"Collection data", ~"Disease inputs", ~"ICD-10 codes", ~"Process", 
"Chronic obstructive pulmonary disease", "Population and mortality rates (TO DO: Belen check)", "Incidence, prevalence and mortality (@RN8)", "J41-J42.4, J43-J44.9", "Remission set to zero", 
"Diabetes mellitus type 2", "Population and mortality rates (TO DO: Belen check)", "Incidence, prevalence and mortality (@RN8)", "E08-E08.11, E08.3-E08.9, E12-E12.1, E12.3-E13.11, E13.3-E14.1, E14.3-E14.9, R73-R73.9 (all diabetes)", "Remission set to zero",
"Ischemic heart disease", "Population and mortality rates (TO DO: Belen check)", "Incidence, prevalence and mortality (@RN8)", "I20-I21.6, I21.9-I25.9, Z82.4-Z82.49", "Remission set to zero and higher weight to prevalence (half bar)", 
"Stroke", "Population and mortality rates (TO DO: Belen check)", "Incidence, prevalence and mortality (@RN8)", "G45-G46.8, I60-I62, I62.9-I64, I64.1, I65-I69.998", "Remission set to zero and higher weight to prevalence input (half bar)",
"Breast cancer (females)", "Population and deaths data for Australia (@RN1)", "Incidence and mortality (@RN3)", "C50", "Remission set to zero", 
"Uterine cancer (females)",  "Population and deaths data for Australia (@RN1)", "Incidence and mortality (@RN3)", "C54–C55", "Remission set to zero", 
"Colon cancer",  "Population and deaths data for Australia (@RN1)", "Incidence and mortality (@RN3)", "C18", "Remission set to zero", 
"Lung cancer",  "Population and deaths data for Australia (@RN1)", "Incidence and mortality (@RN3)", "C33–C34", "Remission set to zero"
)

kable(dismod_process , booktabs = T, caption = "Dismod II input data and procedures") %>%
 kable_styling(latex_options = c("hold position", "repeat_header"),
               full_width = F)%>%
  column_spec(1, width = "5em") %>%
  column_spec(2:5, width = "8em") %>%
  row_spec(0, bold = T)

```

#### Calculate diseases trends 

Table 3 outlines methods used to derive all-cause mortality (TO DO), diseases and injuries trends (TO DO) followed by the code to obtain diseases trends from the source data (*calculateDiseaseTrends*). Trends are used in the main code to run the model (*HERE*) modifying corresponding inptus. 

```{r,echo=FALSE, message=FALSE, warning=FALSE}

trends_diseases <- tribble(
~"Input", ~"Inputs", ~"Assumptions", ~"Data processing",
"General life table: all cause mortality", "TO DO", "1. Age-standardised trends by sex applied to all groups, and 2. Number of years trends applied into the future equivalent to the original trends forecast (e.g. if from 2017 to 2020, then trends were applied for 3 years and then stabilised", "TO DO", 
"Disease life table: incidence cancers", "@RN3 forecast from 2017 to 2020", "Same above", "We derive an annual trend as the log difference between the age standardised rate in 2020 and 2017 divided by the diffence in years",
"Disease life table: case fatality cancers", "@RN4 forecast from 2014 to 20
for mortality", "Same above", "Same above",
"Disease life table: cardiovascular diseases incidence", "@RN5 hospitalization past trends (actual data) from years 2001 to 2018", "Same above and forecast trends were applied to ischemic heart disease and stroke", "Future Age standardized rates (ASR) were freest derived from ASR from past trends by sex using linear regression for predominately increasing trends and log-linear regression for decreasing trends. Trends were estimated up to year 2023. Annual trends were derived as above (ln(ASR2023/ASR2018)/years)", 
"Disease life table: cardiovascular disease case fatality", "@RN5 mortality past trends (actual data) from years 2001 to 2018", "Same above and forecast trends were applied to ischemic heart disease and stroke", "Future Age standardized rates (ASR) were first derived from ASR from past trends by sex using linear regression for predominately increasing trends and log-linear regression for decreasing trends. Trends were estimated up to year 2023. Annual trends were derived as above (ln(ASR2023/ASR2018)/years)",
"Disease life table: diabetes incidence", "@RN6 mortality past trends from years 1997 to 2023 for diabetes as the underlying or associated cause of death", "Same as above and trends only applied to incidence", "Same as above with forecast up to year 2023", 
"Disease life table: Chronic obstructive pulmonary disease incidence", "@RN7 mortality past trends from year 2005 to 2018", "Same as above", "Same as above"
)

kable(trends_diseases, booktabs = T, caption = "Forecast trends for all-cause mortality, diseases and injuries") %>%
 kable_styling(latex_options = c("repeat_header", "hold position"),
               full_width = F)%>%
  column_spec(1:3, width = "8em") %>%
  column_spec(4, width = "14em") %>%
  row_spec(0, bold = T)
```

```{r,echo=FALSE, message=FALSE}

source("Scripts/data_prep/trends_prep.R")
options(scipen=999)
  trends_diseases <- calculateDiseaseTrends(
 incidence_trends_cancers="Data/original/aihw/cancer_incidence_AIHW_with_projections.xlsx",
  mortality_trends_cancers="Data/original/aihw/cancers_trends_mortality_aihw.xls",
  trends_cvd="Data/original/aihw/cardiovascular_disease_trends_aihw.xlsx",
  grim_books="Data/original/aihw/grim_books_utf8.csv",
  trends_diabetes="Data/original/aihw/diabetes_trends_aihw.xls"
)

write.csv(trends_diseases[[1]], "Data/processed/mslt/mortality_trends_f.csv", row.names=F, quote=T)
write.csv(trends_diseases[[2]], "Data/processed/mslt/mortality_trends_m.csv", row.names=F, quote=T)
write.csv(trends_diseases[[3]], "Data/processed/mslt/incidence_trends_f.csv", row.names=F, quote=T)
write.csv(trends_diseases[[4]], "Data/processed/mslt/incidence_trends_m.csv", row.names=F, quote=T)
```

### Generate multi-state life table dataframe

Finally, based on the above data preparation, we use *CalculateMSLT* to generate the input dataframe to run the PMSLT code (*HERE*).

```{r,echo=FALSE, message=FALSE}
source("Scripts/data_prep/mslt_gbd_prep.R")
options(scipen=999)
mslt <- calculateMSLT(
  gbd_wider="Data/processed/mslt/gbd_wider.csv",
  dismod_output_cancers="Data/processed/mslt/dismod_output_cancers.csv",
  dismod_output_non_cancers="Data/processed/mslt/dismod_output_non_cancers.csv")
write.csv(mslt, "Data/processed/mslt/mslt_df.csv", row.names=F, quote=T)
```

